Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-04 ms.tex
intro

A better understanding of what mechanisms regulate accretion onto
quiescent SMBHs would have implications for a variety of observations,
such as constraints on the occupation fraction of SMBHs in low mass
galaxies.  \citet{Miller+15} use the average relationship between the
nuclear X-ray luminosities, $L_{X}$, of a sample of elliptical
galaxies and their associated SMBH masses to infer tentative evidence for a
SMBH occupation fraction less than unity in galaxies with stellar
masses $M_{\star} \lesssim 10^{10}M_{\odot}$ ($M_{\bullet} \lesssim
10^{7}M_{\odot}$).  This method relies on extrapolating a power-law
fit of the $L_X-\Mbh$ distribution to low values of $L_{\rm X}$ below
the detection threshold.  This assumption is questionable if
different physical processes control the accretion rates onto the
lowest mass SMBHs.

@@

The gas density in galactic nuclei also influences the emission from
stellar tidal disruption events (TDEs), such as the high energy
transient {\it Swift} J1644+57 (\citealt{Levan+11};
\citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).
This transient event was powered by an impulsive relativistic jet,
which produced synchrotron radio emission as the jet material was
decelerated by shock interaction with the CNM of the previously
quiescent SMBH (\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).
Detailed modeling of J1644+57 showed that the CNM density was much
lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{BergerZauderer+:2012a}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in J1644+57.
Variations in the properties of the CNM could help explain why most
TDEs appear to be radio quiet (e.g.~\citealt{Bower+13};
\citealt{VanVelzen+13}).

@@

\citet{Ho:2009a} determines the SMBH accretion rates in a sample of
early-type galaxies using X-ray observations to determine the
Bondi accretion rate and based on estimates for the mass loss rate of
evolved stars.  Both methods lead to the conclusion that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Much
evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to the low efficiency with which the inflowing gas ultimately reaches the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Past studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  A key finding of these works is the periodic
development of cooling instabilities on galactic scales, which
temporarily increase the gas accretion rate towards the nucleus until
feedback becomes effective to shut off the flow and halt
SMBH growth.  Here we instead focus on time-independent models,
neglecting radiative cooling.  This approach allows us to
systematically explore the relevant parameter space and to derive
simple analytic expressions that prove useful in exploring whether
cooling instabilities manifest across the expected range of galaxy
properties, or whether other (non-AGN) forms of feedback can result in
steady accretion.  Even if cooling instabilities develop on galactic
scales over longer $\sim$ Gyr timescales, a state of
quasi-steady accretion may exist between these inflow events on
smaller radial scales comparable to the sphere of influence.  We are
thus motivated to explore whether steady flow is achieved within the
nucleus under any circumstances, or whether the gas content is always
building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases resides near the nominal Bondi
radius).  When heating is sufficiently weak, however, the stagnation
radius may not exist at all, i.e. the entire ISM of the galaxy is flowing
towards the SMBH.  A goal of this work is to assess under what
conditions a stagnation radius exists, as this question is intimately
tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

>>>>

A better understanding of what mechanisms regulate accretion onto
quiescent SMBHs would have implications for a variety of observations,
such as constraints on the occupation fraction of SMBHs in low mass
galaxies.  \citet{Miller+15} use the average relationship between the
nuclear X-ray luminosities, $L_{X}$, of a sample of elliptical
galaxies and their associated SMBH masses to infer tentative evidence for a
SMBH occupation fraction less than unity in galaxies with stellar
masses $M_{\star} \lesssim 10^{10}M_{\odot}$ ($M_{\bullet} \lesssim
10^{7}M_{\odot}$).  This method relies on extrapolating a power-law
fit of the $L_X-\Mbh$ distribution to low values of $L_{\rm X}$ below
the instrument detection threshold, an assumption which is questionable if
different physical processes control the accretion rates onto the
lowest mass SMBHs.

@@

The gas density in galactic nuclei also influences the emission from
stellar tidal disruption events (TDEs), such as the high energy
transient {\it Swift} J1644+57 (\citealt{Levan+11};
\citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).
This event was powered by an impulsive relativistic jet,
which produced synchrotron radio emission as the jet material was
decelerated by shock interaction with the CNM of the previously
quiescent SMBH (\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).
Detailed modeling of J1644+57 showed that the CNM density was much
lower than that measured surrounding Sgr A$^{\star}$ on a similar radial scale
(\citealt{Metzger+12}; \citealt{BergerZauderer+:2012a}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing potentially much different radio emission than in J1644+57.
Variations in the properties of the CNM could help explain why most
TDEs appear to be radio quiet (e.g.~\citealt{Bower+13};
\citealt{VanVelzen+13}).

@@

\citet{Ho:2009a} determines the SMBH accretion rates in a sample of
early-type galaxies using X-ray observations to determine the
Bondi accretion rate and based on estimates for the mass loss rate of
evolved stars.  Both methods lead to the conclusion that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Several lines of evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to the low efficiency with which the inflowing gas ultimately reaches the SMBH due to disk outflows (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Past studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  A key finding of these works is the periodic
development of cooling instabilities on galactic scales, which
temporarily increase the gas accretion rate towards the nucleus until
feedback becomes effective to shut off the flow and halt
SMBH growth.  Here we instead focus on time-independent models,
neglecting radiative cooling.  This approach allows us to
systematically explore the relevant parameter space and to derive
simple analytic expressions that prove useful to determine whether
cooling instabilities manifest across the expected range of galaxy
properties, or whether other (non-AGN) forms of feedback can result in
steady accretion.  Even if cooling instabilities develop on galactic
scales over longer $\sim$ Gyr timescales, a state of
quasi-steady accretion may exist between these inflow events on
smaller radial scales comparable to the sphere of influence.  We are
thus motivated to explore whether steady flow is typically expected within the
nucleus under, or whether in most galaxies the nuclear gas content is always
building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ will in many cases reside near the nominal Bondi
radius).  When heating is sufficiently weak, however, the stagnation
radius may move to much larger radii or not exist at all, in which case the entire ISM of the galaxy is flowing
towards the SMBH.  A goal of this work is to assess under what
conditions a stagnation radius exists, as this question is intimately
tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.
